### 2021-08-27 ms/ms.Rmd
wrote README
added supporting information files

We used the R packages **taxonlookup** version `r packageVersion("taxonlookup")` [@pennell_simple_2016] and **V.phylomaker** version `r packageVersion("V.phylomaker")` [@jin_vphylomaker_2019] to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018 ; @zanne_three_2014]. We further resolved large ($\ge4$ taxa) polytomies in `r nrow(polytomies) - 1` clades with sufficient sequence data using **PyPHLAWD** version 1.0 [@smith_pyphlawd_2019] in `r system("python3 --version", intern = TRUE)` (Python Software Foundation, https://www.python.org/). We used sequence data from the most recent GenBank Plant and Fungal sequences database division [@ouellette_database_1997]. We inferred subtree phylogenies using `r str_extract(str_c(system("raxmlHPC-PTHREADS -version", intern = TRUE), collapse = ""), "RAxML version [0-9]*.[0-9]*.[0-9]*")` [@stamatakis_raxml_2014] and conducated molecular dating using the `chronos()` function in the R package **ape** version `r packageVersion("ape")` [@paradis_ape_2019] to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for **PyPHLAWD**. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the `multi2di()` function in **ape**. The final data set for which we had both trait and phylogenetic information contained `r n_tax3` taxa. Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

We extracted `r n_pair` phylogenetically independent taxon pairs. A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: $\tt{((A,B),(C,D))}$ or $\tt{((A,B),C),D))}$. Taxon pairs include all comparisons of $\tt{A}$ with $\tt{B}$ and $\tt{C}$ with $\tt{D}$ in each four-taxon clade. We extracted pairs using the `extract_sisters()` function in R package **diverge** version `r packageVersion("diverge")` [@anderson_diverge_2021] and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts [@felsenstein_phylogenies_1985] as the difference in the log$_{10}$-transformed trait value. Contrasts are denoted as $\Delta \text{log(trait)}$. We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

Stomatal length negatively covaries with stomatal density similarly on both surfaces, but on the adaxial surface there are many more taxa that have low stomatal density and small size compared to the abaxial surface (Fig. \ref{fig:h1_raw}). In principle, this pattern could arise either because size-density covariance differs or the variance in adaxial stomatal density increases faster than that for abaxial stomatal density. Also note that the interspecific variance increases with time since divergence for all traits (Table \ref{table:model_output}), but the covariance matrix structure did not change qualitatively over time (results not shown). For consistency, we therefore report estimates conditional on time since divergence set to 0. Across pairs, we estimate that the covariance between size and density is similar. The median estimate is $\text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})] - \text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] = `r dplyr::filter(h1_parameters, name == "diff_cov")[["estimate"]]`$, but 0 is within the range of uncertainty (`r dplyr::filter(h1_parameters, name == "diff_cov")[["interval"]]`). However the variance in adaxial stomatal density is significantly greater than the abaxial stomatal density [Fig. \ref{fig:h1}]). We estimate $\text{Var}[\Delta \text{log}(D_\text{ad})]$ is $`r dplyr::filter(h1_parameters, name == "diff_sigma2_d")[["estimate"]]`$ (`r dplyr::filter(h1_parameters, name == "diff_sigma2_d")[["interval"]]`) greater than $\text{Var}[\Delta \text{log}(D_\text{ab})]$. The variance in stomatal length was similar for both surfaces, with an estimate of $`r dplyr::filter(h1_parameters, name == "diff_sigma2_l")[["estimate"]]`$ (`r dplyr::filter(h1_parameters, name == "diff_sigma2_l")[["interval"]]`).

@@

CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data. We thank Jacob Watts for comments on earlier versions of this manuscript.

@@

@@

```{r}

# write supporting information table
write_csv(pair_div, "ms/tableS2.csv")

>>>>

We used the R packages **taxonlookup** version `r packageVersion("taxonlookup")` [@pennell_simple_2016] and **V.phylomaker** version `r packageVersion("V.phylomaker")` [@jin_vphylomaker_2019] to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018 ; @zanne_three_2014]. We further resolved large ($\ge4$ taxa) polytomies in `r nrow(polytomies) - 1` clades with sufficient sequence data using **PyPHLAWD** version 1.0 [@smith_pyphlawd_2019] in `r system("python3 --version", intern = TRUE)` (Python Software Foundation, https://www.python.org/). We used sequence data from the most recent GenBank Plant and Fungal sequences database division [@ouellette_database_1997]. We inferred subtree phylogenies using `r str_extract(str_c(system("raxmlHPC-PTHREADS -version", intern = TRUE), collapse = ""), "RAxML version [0-9]*.[0-9]*.[0-9]*")` [@stamatakis_raxml_2014] and conducated molecular dating using the `chronos()` function in the R package **ape** version `r packageVersion("ape")` [@paradis_ape_2019] to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for **PyPHLAWD**. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the `multi2di()` function in **ape**. The final data set for which we had both trait and phylogenetic information contained `r n_tax3` taxa (Notes S1). Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

We extracted `r n_pair` phylogenetically independent taxon pairs (Table \@ref(tab:pair_div)). A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: $\tt{((A,B),(C,D))}$ or $\tt{((A,B),C),D))}$. Taxon pairs include all comparisons of $\tt{A}$ with $\tt{B}$ and $\tt{C}$ with $\tt{D}$ in each four-taxon clade. We extracted pairs using the `extract_sisters()` function in R package **diverge** version `r packageVersion("diverge")` [@anderson_diverge_2021] and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts [@felsenstein_phylogenies_1985] as the difference in the log$_{10}$-transformed trait value. Contrasts are denoted as $\Delta \text{log(trait)}$. We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

Stomatal length negatively covaries with stomatal density similarly on both surfaces, but on the adaxial surface there are many more taxa that have low stomatal density and small size compared to the abaxial surface (Fig. \ref{fig:h1_raw}). In principle, this pattern could arise either because size-density covariance differs or the variance in adaxial stomatal density increases faster than that for abaxial stomatal density. Also note that the interspecific variance increases with time since divergence for all traits (Table \ref{tab:modeloutput}), but the covariance matrix structure did not change qualitatively over time (results not shown). For consistency, we therefore report estimates conditional on time since divergence set to 0. Across pairs, we estimate that the covariance between size and density is similar. The median estimate is $\text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})] - \text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] = `r dplyr::filter(h1_parameters, name == "diff_cov")[["estimate"]]`$, but 0 is within the range of uncertainty (`r dplyr::filter(h1_parameters, name == "diff_cov")[["interval"]]`). However the variance in adaxial stomatal density is significantly greater than the abaxial stomatal density [Fig. \ref{fig:h1}]). We estimate $\text{Var}[\Delta \text{log}(D_\text{ad})]$ is $`r dplyr::filter(h1_parameters, name == "diff_sigma2_d")[["estimate"]]`$ (`r dplyr::filter(h1_parameters, name == "diff_sigma2_d")[["interval"]]`) greater than $\text{Var}[\Delta \text{log}(D_\text{ab})]$. The variance in stomatal length was similar for both surfaces, with an estimate of $`r dplyr::filter(h1_parameters, name == "diff_sigma2_l")[["estimate"]]`$ (`r dplyr::filter(h1_parameters, name == "diff_sigma2_l")[["interval"]]`).

@@

We thank Jacob Watts for comments on earlier versions of this manuscript.

# Author Contribution

CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data.

@@

**Note S1:** Final fully resolved bifurcating phylogenetic tree of `r n_tax3` taxa in Newick format.
```{r}
library(ape)
write.tree(trimmed_phy, "ms/notesS1.new")
```

@@

\begin{table}
\caption{\label{tab:pair_div}Final data set of `r n_pair` taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny (Note S1). $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

```{r}
# write supporting information table
write_csv(pair_div, "ms/tableS2.csv")
```

```{r, modeloutput}

# stomata_anatomy_metadata |>
#   dplyr::mutate(Source = stringr::str_c("@", source), Taxa = taxa) |>
#   dplyr::select(Source, Taxa) |>
#   knitr::kable(format = "markdown", caption = "Primary sources of stomatal anatomical data and the taxa covered by each source.")
tibble(x = 1:2, y = 3:4) |>
  knitr::kable(format = "markdown", caption = "Model output")
```

### 2021-08-27 ms/ms.tex
wrote README
added supporting information files

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018} ; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 30 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{S. A. Smith and Walker 2019}) in Python 3.8.2 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducated molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.5 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 711 taxa. Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

We extracted 327 phylogenetically independent taxon pairs. A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: \(\tt{((A,B),(C,D))}\) or \(\tt{((A,B),C),D))}\). Taxon pairs include all comparisons of \(\tt{A}\) with \(\tt{B}\) and \(\tt{C}\) with \(\tt{D}\) in each four-taxon clade. We extracted pairs using the \texttt{extract\_sisters()} function in R package \textbf{diverge} version 2.0.1 (\protect\hyperlink{ref-anderson_diverge_2021}{Anderson and Weir 2021}) and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts (\protect\hyperlink{ref-felsenstein_phylogenies_1985}{Felsenstein 1985}) as the difference in the log\(_{10}\)-transformed trait value. Contrasts are denoted as \(\Delta \text{log(trait)}\). We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

Stomatal length negatively covaries with stomatal density similarly on both surfaces, but on the adaxial surface there are many more taxa that have low stomatal density and small size compared to the abaxial surface (Fig. \ref{fig:h1_raw}). In principle, this pattern could arise either because size-density covariance differs or the variance in adaxial stomatal density increases faster than that for abaxial stomatal density. Also note that the interspecific variance increases with time since divergence for all traits (Table \ref{table:model_output}), but the covariance matrix structure did not change qualitatively over time (results not shown). For consistency, we therefore report estimates conditional on time since divergence set to 0. Across pairs, we estimate that the covariance between size and density is similar. The median estimate is \(\text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})] - \text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] = 2.66 \times 10^{-3}\), but 0 is within the range of uncertainty (95\% HPD interval \([-6.72 \times 10^{-4},6.79 \times 10^{-3}]\)). However the variance in adaxial stomatal density is significantly greater than the abaxial stomatal density {[}Fig. \ref{fig:h1}{]}). We estimate \(\text{Var}[\Delta \text{log}(D_\text{ad})]\) is \(2.64 \times 10^{-2}\) (95\% HPD interval \([8.40 \times 10^{-3},5.13 \times 10^{-2}]\)) greater than \(\text{Var}[\Delta \text{log}(D_\text{ab})]\). The variance in stomatal length was similar for both surfaces, with an estimate of \(-8.08 \times 10^{-5}\) (95\% HPD interval \([-1.38 \times 10^{-3},1.23 \times 10^{-3}]\)).

@@

CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data. We thank Jacob Watts for comments on earlier versions of this manuscript.

@@

@@

>>>>

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018} ; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 30 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{S. A. Smith and Walker 2019}) in Python 3.8.2 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducated molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.5 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 711 taxa (Notes S1). Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

We extracted 327 phylogenetically independent taxon pairs (Table @ref(tab:pair\_div)). A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: \(\tt{((A,B),(C,D))}\) or \(\tt{((A,B),C),D))}\). Taxon pairs include all comparisons of \(\tt{A}\) with \(\tt{B}\) and \(\tt{C}\) with \(\tt{D}\) in each four-taxon clade. We extracted pairs using the \texttt{extract\_sisters()} function in R package \textbf{diverge} version 2.0.1 (\protect\hyperlink{ref-anderson_diverge_2021}{Anderson and Weir 2021}) and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts (\protect\hyperlink{ref-felsenstein_phylogenies_1985}{Felsenstein 1985}) as the difference in the log\(_{10}\)-transformed trait value. Contrasts are denoted as \(\Delta \text{log(trait)}\). We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

Stomatal length negatively covaries with stomatal density similarly on both surfaces, but on the adaxial surface there are many more taxa that have low stomatal density and small size compared to the abaxial surface (Fig. \ref{fig:h1_raw}). In principle, this pattern could arise either because size-density covariance differs or the variance in adaxial stomatal density increases faster than that for abaxial stomatal density. Also note that the interspecific variance increases with time since divergence for all traits (Table \ref{tab:modeloutput}), but the covariance matrix structure did not change qualitatively over time (results not shown). For consistency, we therefore report estimates conditional on time since divergence set to 0. Across pairs, we estimate that the covariance between size and density is similar. The median estimate is \(\text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})] - \text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] = 2.66 \times 10^{-3}\), but 0 is within the range of uncertainty (95\% HPD interval \([-6.72 \times 10^{-4},6.79 \times 10^{-3}]\)). However the variance in adaxial stomatal density is significantly greater than the abaxial stomatal density {[}Fig. \ref{fig:h1}{]}). We estimate \(\text{Var}[\Delta \text{log}(D_\text{ad})]\) is \(2.64 \times 10^{-2}\) (95\% HPD interval \([8.40 \times 10^{-3},5.13 \times 10^{-2}]\)) greater than \(\text{Var}[\Delta \text{log}(D_\text{ab})]\). The variance in stomatal length was similar for both surfaces, with an estimate of \(-8.08 \times 10^{-5}\) (95\% HPD interval \([-1.38 \times 10^{-3},1.23 \times 10^{-3}]\)).

@@

We thank Jacob Watts for comments on earlier versions of this manuscript.

\hypertarget{author-contribution}{%
\section{Author Contribution}\label{author-contribution}}

CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data.

@@

\textbf{Note S1:} Final fully resolved bifurcating phylogenetic tree of 711 taxa in Newick format.

@@

\begin{table}
\caption{\label{tab:pair_div}Final data set of 327 taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny (Note S1). $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

\begin{longtable}[]{@{}rr@{}}
\caption{\label{tab:modeloutput}Model output}\tabularnewline
\toprule
x & y \\
\midrule
\endfirsthead
\toprule
x & y \\
\midrule
\endhead
1 & 3 \\
2 & 4 \\
\bottomrule
\end{longtable}
